---
title: "Project 2"
author: "Darren Cohen, Andrew Gutierrez"
date: "May 10, 2018"
output: html_document
---

<!-- change echo=FALSE to echo=TRUE to show code -->
```{r global_options, include=FALSE}
knitr::opts_chunk$set(prompt=TRUE, comment="", echo=TRUE)
```

For our project we will be using drug death data to see if there is correlation between aspects of a subject's life and their drug usage. We will also be looking into the kinds of drugs that were present in the subject's system at the time of their death. Another aspect we are interested in looking at is where the person took the drugs compared to where they died, as the data has an attribute showing where they took the drug and another showing their location when they passed away.

```{r}
dat=read.csv("https://raw.githubusercontent.com/AndrewGutierrez/drugdeaths/master/Accidental_Drug_Related_Deaths__2012-2017.csv")
```

***

###Data Preprocessing

Before we can start working with the data, we need to clean and process it. To begin we will remove some of the rows that will not be useful for our project. This will include the 'CaseNumber', 'AmendedMannerofDeath', 'DescriptionofInjury', and 'MannerofDeath'. The 'CaseNumber' column will be removed as all values in this are unique, as the are used to identify each row. The 'AmendedMannerofDeath' and 'MannerofDeath' columns both consist of the same values for each row, thus provding information that is not needed. The 'DescriptionofInjury' column will be removed because most of the values are missing, and the values that are there are mostly unique so we can not fill in the missing values.

```{r}
dat$CaseNumber=NULL
dat$AmendedMannerofDeath=NULL
dat$MannerofDeath=NULL
dat$DescriptionofInjury=NULL
```

Another step we need to take is to see if we have any NA values in our data set. We can do this by looking at the number of rows in the data set minus the number of complete cases.

```{r}
nrow(dat)-sum(complete.cases(dat))
```

After seeing that there are incomplete cases in our data set, we need to further identify which column is causing these three incomplete cases. A way this can be done is by using the 'summmary()' function.

```{r}
summary(dat)
```

From the summary we see that the 'Age' column has these three NA values, to fix this we will be replacing these values with the average age from our data set.

```{r}
avg_age=round(mean(dat[!is.na(dat$Age),]$Age))
dat[is.na(dat$Age),]$Age =avg_age
```

Looking at the summary, we can also see if there are other aspects of our data set that need to be refined. From looking at the 'Race' column we saw that there were some rows that contained empty strings. There were also rows that contained the value "Unkown", because of this we decided to convert the empty string values into "Unknown". We will also be converting the empty strings in 'Sex' into either "Male" or "Female" by randomly picking one of the two genders. Looking at the 'Location' column, there were three main values: "Residence", "Hospital", and "Other", the remaining values in this column will be converted into "Other".

```{r}
#not sure why running this code twice gives an error
dat$Race=ifelse(dat$Race=="", "Unknown", levels(dat$Race)[dat$Race])

dat$Sex=ifelse(dat$Sex=="", sample(c("Male", "Female"), 1, replace=TRUE), levels(dat$Sex)[dat$Sex])

dat$Location=ifelse(dat$Location!="Residence" & dat$Location!="Hospital", "Other", levels(dat$Location)[dat$Location])

dat$InjuryPlace=ifelse(dat$InjuryPlace=="", "Unknown", levels(dat$InjuryPlace)[dat$InjuryPlace])
```

Another thing we want to do is make additional features that we can use when using machine learning to make predictions. These features will have "Yes" or "No" values based on if another feature matches a classification we are looking for. Along with creating these additional features, the type of drug detected is marked in the respected column with a "Y" or "y" meaning the drug was present, and a value of an empty string if the drug was not present. We will convert all "y" values into "Y" so that the indication of yes will match. The empty string values will be converted into "N" to indicate that the drug was not present.  

```{r}
hospital_death=ifelse(dat$Location=="Hospital", "Yes", "No")
dat$hospital=hospital_death

Residence_death=ifelse(dat$Location=="Residence", "Yes", "No")
dat$Residence=Residence_death

heroin=ifelse(dat$Heroin=="Y" | dat$Heroin=="y", "Y", "N")
dat$Heroin=heroin

cocaine=ifelse(dat$Cocaine=="Y" | dat$Cocaine=="y", "Y", "N")
dat$Cocaine=cocaine

Oxycodone=ifelse(dat$Oxycodone=="Y" | dat$Oxycodone=="y", "Y", "N")
dat$Oxycodone=Oxycodone

Oxymorphone=ifelse(dat$Oxymorphone=="Y" | dat$Oxymorphone=="y", "Y", "N")
dat$Oxymorphone=Oxymorphone

EtOH=ifelse(dat$EtOH=="Y" | dat$EtOH=="y", "Y", "N")
dat$EtOH=EtOH

Hydrocodone=ifelse(dat$Hydrocodone=="Y" | dat$Hydrocodone=="y", "Y", "N")
dat$Hydrocodone=Hydrocodone

Amphet=ifelse(dat$Amphet=="Y" | dat$Amphet=="y", "Y", "N")
dat$Amphet=Amphet

Methadone=ifelse(dat$Methadone=="Y" | dat$Methadone=="y", "Y", "N")
dat$Methadone=Methadone

Tramad=ifelse(dat$Tramad=="Y" | dat$Tramad=="y", "Y", "N")
dat$Tramad=Tramad

Benzodiazepine=ifelse(dat$Benzodiazepine=="Y" | dat$Benzodiazepine=="y" | dat$Benzodiazepine==" Y", "Y", "N")
dat$Benzodiazepine=Benzodiazepine

Fentanyl=ifelse(dat$Fentanyl=="", "N", "Y")
dat$Fentanyl=Fentanyl
```

***

###Data Exploration

Before we use our data set to make predictions, we need to explore the data to see what trends we can see or to find any aspects of the data we may want to explore further.

```{r}
plot(density(dat$Age), main="Density of age in the data set", col="cadetblue")
grid()
```

From making a density plot of the ages in the data set, we see that there is a good range of ages from 20 to 65. From this we also see that our average for ages will be around 42.

```{r}
dat$years=substr(dat$Date, 7, nchar(dat$Date))
barplot(table(dat$years), main="Number of deaths per year", col="coral")
```

If we take a look at a barplot of how often the years appear in the data set, we see that there is an upward trend in the number of deaths per year. This is something that we may want to exxplore further, as the number of deaths per year due to drug deaths is increasing. 

```{r}
par(mar=c(3,10,3,3))
barplot(sort(table(dat$Race)), main="Number of deaths per race", col="coral", las=1, horiz=TRUE)
```

By looking at the number of deaths per race in the data set, we see if race may be something we would want to include as a feature when making predictions in our models.

```{r}
barplot(table(dat$Sex), main="Deaths per sex", col="coral", ylim=c(0,3000))
```

By looking at a barplot of the number of deaths per sex, we see that there are about three times as many males in the data set as there are females.

```{r}
table(dat$InjuryPlace)
```

By looking at a table of where a subject took the drugs, we see that there is a great varity of places listed. If we want to be able to use this attribute for making predictions we may want to bin this data as we did for where a person died. 

```{r}
table(dat$Location)
```

Building off of tthe table where a subject took the drugs, we have a table showing where the subjects died. This could be a great attribute to use in determining which types of drugs lead to subjects getting some kind of treatment to try and save their life. 

```{r}
tbl=sapply(dat[, 14:24], table)
par(mar=c(3,8,3,3))
barplot(sort(tbl[2,]), main="Number of recorded uses per drug type", las=1, horiz=TRUE, col="coral", xlim=c(0, 2500))
```

Since the data set contains various columns showing if a certain drug present in the subject's system, we can use this to see which kinds of drugs are being used. The plot shows that there are five different drugs used in our data set of about 4100 rows. This means that there will be subjects that had more than one drug at a time in their system. This could be useful in looking into if combination of drugs contribute to a person's death more than just those drugs by themselves.

***

###Prediction

```{R}
dat$years=as.numeric(dat$years)
fit=glm(years ~ Age, data=dat)

```

```{r}
probs = predict(fit, newdata=dat, type="response")
probs=round(probs)
```

```{r}
library(rpart)
library(rpart.plot)

```

```{r}
fit = rpart(Sex ~ Age, data=dat, method="class")
```






